---
title: "Ancestral State Reconstruction with coevolve"
format:
  html:
    toc: true
    code-fold: false
    fig-width: 8
    fig-height: 6
execute:
  warning: false
  message: false
---

## Overview

The `coev_ancestral_states()` function extracts posterior estimates of latent
trait values at internal (ancestral) nodes of the phylogeny from a fitted
`coevfit` model. Because the model estimates latent states at every node in the
tree, ancestral state reconstruction is a natural byproduct of model fitting ---
no separate analysis is needed.

This notebook demonstrates the function using the **authority** dataset bundled
with the package: political and religious authority among 97 Austronesian
societies.

## Setup

```{r}
#| label: setup
library(coevolve)
library(ggplot2)
library(dplyr)
library(tidyr)
library(ape)
theme_set(theme_minimal(base_size = 13))
```

## Fit the model

We fit a coevolutionary model with two ordered-logistic variables. For
demonstration purposes we use moderate sampling settings; in practice you would
want more iterations and to check convergence carefully.

```{r}
#| label: fit-model
#| cache: true
fit <- coev_fit(
  data = authority$data,
  variables = list(
    political_authority = "ordered_logistic",
    religious_authority = "ordered_logistic"
  ),
  id = "language",
  tree = authority$phylogeny,
  parallel_chains = 4,
  chains = 4,
  iter_sampling = 500,
  iter_warmup = 500,
  seed = 42
)
```

Quick check that the model sampled adequately:

```{r}
#| label: model-summary
summary(fit)
```

## Ancestral states on the latent scale

The default call returns posterior summaries (median + 95% credible interval) for
all internal nodes on the latent (eta) scale:

```{r}
#| label: asr-latent
asr_latent <- coev_ancestral_states(fit)
head(asr_latent)
```

The result is a tibble with one row per node-variable combination. The `node`
column uses ape's numbering convention (tips are `1:N_tips`, internal nodes are
`(N_tips+1):(2*N_tips-1)`), so it integrates directly with phylo objects.

### Visualizing latent ancestral states on the tree

We can paint the tree with estimated ancestral values by mapping the posterior
median to a diverging color gradient. Blue indicates low latent values; red
indicates high:

```{r}
#| label: plot-latent-tree
#| fig-height: 10
tree <- attr(asr_latent, "ref_tree")
n_tips <- length(tree$tip.label)
asr_all <- coev_ancestral_states(fit, nodes = "all")

pal <- colorRampPalette(c("#2166AC", "#F7F7F7", "#B2182B"))(100)
global_range <- range(asr_all$estimate)

par(mfrow = c(1, 2), mar = c(2, 0, 2, 0))

for (var in unique(asr_all$variable)) {
  df_var <- asr_all |> filter(variable == var)
  vals <- setNames(df_var$estimate, df_var$node)

  plot(tree, show.tip.label = FALSE, edge.width = 1.5,
       edge.color = "grey70", main = gsub("_", " ", var),
       no.margin = FALSE)

  internal_ids <- (n_tips + 1):(n_tips + tree$Nnode)
  internal_vals <- vals[as.character(internal_ids)]
  bins <- as.integer(
    cut(internal_vals,
        breaks = seq(global_range[1], global_range[2], length.out = 101))
  )
  bins[is.na(bins)] <- 50
  bins <- pmax(1, pmin(100, bins))
  nodelabels(pch = 21, bg = pal[bins], col = NA, cex = 1.2)
}

# add shared color bar legend
par(fig = c(0.35, 0.65, 0.0, 0.05), new = TRUE, mar = c(0, 0, 0, 0))
image(
  x = seq(global_range[1], global_range[2], length.out = 100),
  y = 1,
  z = matrix(1:100, nrow = 100),
  col = pal,
  axes = FALSE, xlab = "", ylab = ""
)
axis(1, at = round(seq(global_range[1], global_range[2], length.out = 5), 1),
     cex.axis = 0.8, padj = -1.5, tck = -0.3)
mtext("Latent value (eta)", side = 1, line = 1.2, cex = 0.7)
```

### Joint latent-space scatterplot

We can also look at the joint distribution of ancestral states across the two
variables. Each point is an internal node, with crosshairs showing the 95%
credible interval:

```{r}
#| label: plot-latent-scatter
asr_wide <- asr_latent |>
  select(node, variable, estimate, lower, upper) |>
  pivot_wider(
    names_from = variable,
    values_from = c(estimate, lower, upper)
  )

ggplot(asr_wide,
       aes(x = estimate_political_authority,
           y = estimate_religious_authority)) +
  geom_errorbar(
    aes(ymin = lower_religious_authority,
        ymax = upper_religious_authority),
    alpha = 0.2, width = 0
  ) +
  geom_errorbarh(
    aes(xmin = lower_political_authority,
        xmax = upper_political_authority),
    alpha = 0.2, height = 0
  ) +
  geom_point(color = "#B2182B", size = 2) +
  labs(
    x = "Political authority (latent)",
    y = "Religious authority (latent)",
    title = "Ancestral states at internal nodes"
  )
```

## Ancestral states on the response scale

For ordered-logistic variables, the response scale returns estimated category
probabilities at each node --- how likely each level of authority (absent,
sublocal, local, supralocal) was at that ancestral node:

```{r}
#| label: asr-response
asr_resp <- coev_ancestral_states(
  fit, nodes = "all", scale = "response"
)
head(asr_resp)
```

Notice the output now has `prob_1`, `prob_2`, etc. columns instead of
`estimate`/`lower`/`upper`. Each row sums to 1 across probability columns.

### Pie chart tree: response-scale probabilities

This is the most intuitive way to visualize ordinal ancestral states. Each pie
chart at an internal node shows the posterior median probability of each
category. At the tips, observed values are shown as solid colors (one category
with probability 1). If a tip has missing data, the model-estimated
probabilities are shown instead.

```{r}
#| label: plot-pie-tree
#| fig-height: 12
#| fig-width: 10
cat_labels <- c("Absent", "Sublocal", "Local", "Supralocal")
cat_colors <- c("#FFFFB2", "#FECC5C", "#FD8D3C", "#E31A1C")
prob_cols <- grep("^prob_", names(asr_resp), value = TRUE)
n_cats <- length(prob_cols)

# build observed-value lookup: tip label -> integer category per variable
obs_data <- authority$data
obs_data$tip_label <- tree$tip.label[match(obs_data$language,
                                           tree$tip.label)]

par(mfrow = c(1, 2), mar = c(1, 0, 3, 0))

for (var in unique(asr_resp$variable)) {
  df_var <- asr_resp |>
    filter(variable == var) |>
    arrange(node)

  plot(tree, show.tip.label = FALSE, edge.width = 1.5,
       edge.color = "grey50", no.margin = FALSE,
       main = gsub("_", " ", var))

  # internal node pies (model estimates)
  df_internal <- df_var |> filter(node > n_tips) |> arrange(node)
  pie_mat <- as.matrix(df_internal[, prob_cols])
  nodelabels(pie = pie_mat, piecol = cat_colors, cex = 0.6)

  # tip pies: observed values when available, model estimates when missing
  tip_pie <- matrix(0, nrow = n_tips, ncol = n_cats)
  for (i in seq_len(n_tips)) {
    tip_name <- tree$tip.label[i]
    row_idx <- match(tip_name, obs_data$language)
    obs_val <- if (!is.na(row_idx)) obs_data[[var]][row_idx] else NA
    if (!is.na(obs_val)) {
      # observed: deterministic pie (all mass on one category)
      cat_idx <- as.integer(obs_val)
      tip_pie[i, cat_idx] <- 1
    } else {
      # missing: use model-estimated probabilities
      model_row <- df_var |> filter(node == i)
      tip_pie[i, ] <- as.numeric(model_row[, prob_cols])
    }
  }
  tiplabels(pie = tip_pie, piecol = cat_colors, cex = 0.4)
}

# shared legend
par(fig = c(0.3, 0.7, 0.0, 0.05), new = TRUE, mar = c(0, 0, 0, 0))
plot.new()
legend("center", legend = cat_labels, fill = cat_colors,
       horiz = TRUE, bty = "n", cex = 0.9, title = "Authority level")
```

## Working with raw posterior draws

Setting `summary = FALSE` returns the full posterior array, which is useful for
custom summaries or propagating uncertainty:

```{r}
#| label: asr-draws
asr_draws <- coev_ancestral_states(fit, summary = FALSE)
str(asr_draws, max.level = 1)
```

The `draws` element is a 3D array with dimensions `[draws, nodes, variables]`.
Here's an example: the posterior density of latent values at the root node for
both variables:

```{r}
#| label: plot-root-density
root_node_idx <- which(asr_draws$node_ids == n_tips + 1)

root_df <- data.frame(
  political_authority = asr_draws$draws[, root_node_idx, 1],
  religious_authority = asr_draws$draws[, root_node_idx, 2]
) |>
  pivot_longer(everything(), names_to = "variable", values_to = "eta")

ggplot(root_df, aes(x = eta, fill = variable)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(
    values = c(political_authority = "#2166AC",
               religious_authority = "#B2182B"),
    labels = c("Political", "Religious"),
    name = "Authority"
  ) +
  labs(
    x = "Latent value (eta)",
    y = "Density",
    title = "Posterior distribution at the root node"
  )
```

## Identifying nodes on the tree

To query ancestral states for specific nodes, you first need to know their IDs.
The `coev_plot_node_labels()` function plots the tree with internal node IDs
displayed in red:

```{r}
#| label: plot-node-labels
#| fig-height: 10
tree <- coev_plot_node_labels(fit)
```

Node numbering follows ape's convention: tips are `1:N_tips` and internal nodes
are `(N_tips + 1):(2 * N_tips - 1)`.

### Interactive node selection

In an interactive R session you can click directly on the tree to identify nodes.
After plotting, call `identify()` on the returned tree object --- click on nodes
of interest, then press Escape to finish:

```{r}
#| label: identify-nodes
#| eval: false
tree <- coev_plot_node_labels(fit)
selected <- identify(tree)
# selected$nodes contains the node IDs you clicked on
coev_ancestral_states(fit, nodes = selected$nodes)
```

## Selecting specific nodes and variables

Once you know the node IDs, you can target specific nodes or variables to reduce
computation and focus the output:

```{r}
#| label: asr-subset
# only religious authority, at three specific internal nodes
asr_sub <- coev_ancestral_states(
  fit,
  variables = "religious_authority",
  nodes = c(n_tips + 1, n_tips + 2, n_tips + 3)
)
asr_sub
```

## Session info

```{r}
#| label: session-info
sessionInfo()
```
