---
title: "Benchmark: nutpie (JAX) vs Stan"
format:
  html:
    toc: true
    code-fold: true
execute:
  warning: false
  message: false
---

```{r setup}
devtools::load_all()
library(ape)
library(posterior)
library(ggplot2)
library(dplyr)
library(tidyr)

theme_set(theme_minimal(base_size = 14))

# Sampling settings
N_CHAINS  <- 4L
N_WARMUP  <- 500L
N_SAMPLES <- 1000L
SEED      <- 42L
```

## Overview

We benchmark **Stan** (`nuts_sampler = "stan"`) against **nutpie/JAX**
(`nuts_sampler = "nutpie"`) across all vignette models that the JAX backend
supports. For each model and backend we record wall-clock sampling time and
the per-parameter bulk/tail effective sample sizes.

Models:

1. **Authority** — ordered logistic (97 taxa)
2. **Primates** — gamma-log with missing data (21 taxa)
3. **Repeated measures** — normal with repeated obs (20 taxa, 100 obs)
4. **MultiPhylo** — ordered logistic, phylogenetic uncertainty (97 taxa, 2 trees)
5. **Measurement error** — normal with known SE (20 taxa)
6. **Exact GP** — ordered logistic + spatial control via full GP (97 taxa)
7. **HSGP** — ordered logistic + spatial control via Hilbert-space approximate GP (97 taxa, `dist_k=5`)

## Helper functions

```{r helpers}
fit_and_summarise <- function(model_name, ..., nuts_sampler) {
  t0 <- proc.time()[3]
  fit <- coev_fit(
    ...,
    nuts_sampler    = nuts_sampler,
    chains          = N_CHAINS,
    parallel_chains = N_CHAINS,
    iter_warmup     = N_WARMUP,
    iter_sampling   = N_SAMPLES,
    seed            = SEED
  )
  elapsed <- proc.time()[3] - t0

  # Both CmdStanMCMC and jax_fit have $summary()
  summ <- fit$fit$summary()

  data.frame(
    model    = model_name,
    backend  = nuts_sampler,
    variable = summ$variable,
    ess_bulk = summ$ess_bulk,
    ess_tail = summ$ess_tail,
    rhat     = summ$rhat,
    elapsed  = elapsed
  )
}
```

## 1. Authority

```{r authority}
auth_stan <- fit_and_summarise(
  "authority",
  data = authority$data,
  variables = list(
    political_authority = "ordered_logistic",
    religious_authority = "ordered_logistic"
  ),
  id = "language",
  tree = authority$phylogeny,
  prior = list(A_offdiag = "normal(0, 2)"),
  nuts_sampler = "stan"
)

auth_nutpie <- fit_and_summarise(
  "authority",
  data = authority$data,
  variables = list(
    political_authority = "ordered_logistic",
    religious_authority = "ordered_logistic"
  ),
  id = "language",
  tree = authority$phylogeny,
  prior = list(A_offdiag = "normal(0, 2)"),
  nuts_sampler = "nutpie",
  use_grad_based_mass_matrix=FALSE
)
```

## 2. Primates (gamma-log, missing data)

```{r primates}
primates_data <- primates$data[primates$data$clade == "Lemuriformes", ]
primates_phylogeny <- keep.tip(primates$phylogeny, primates_data$species)

prim_stan <- fit_and_summarise(
  "primates",
  data = primates_data,
  variables = list(
    brain_weight = "gamma_log",
    group_size   = "gamma_log"
  ),
  id = "species",
  tree = primates_phylogeny,
  nuts_sampler = "stan"
)

prim_nutpie <- fit_and_summarise(
  "primates",
  data = primates_data,
  variables = list(
    brain_weight = "gamma_log",
    group_size   = "gamma_log"
  ),
  id = "species",
  tree = primates_phylogeny,
  nuts_sampler = "nutpie",
  use_grad_based_mass_matrix=FALSE
)
```

## 3. Repeated measures

```{r repeated}
rep_stan <- fit_and_summarise(
  "repeated",
  data = repeated$data,
  variables = list(x = "normal", y = "normal"),
  id = "species",
  tree = repeated$phylogeny,
  nuts_sampler = "stan"
)

rep_nutpie <- fit_and_summarise(
  "repeated",
  data = repeated$data,
  variables = list(x = "normal", y = "normal"),
  id = "species",
  tree = repeated$phylogeny,
  nuts_sampler = "nutpie",
  use_grad_based_mass_matrix=FALSE
)
```

## 4. MultiPhylo (phylogenetic uncertainty)

```{r multiphylo}
authority_multiphylo <- c(authority$phylogeny, authority$phylogeny)

multi_stan <- fit_and_summarise(
  "multiphylo",
  data = authority$data,
  variables = list(
    political_authority = "ordered_logistic",
    religious_authority = "ordered_logistic"
  ),
  id = "language",
  tree = authority_multiphylo,
  prior = list(A_offdiag = "normal(0, 2)"),
  nuts_sampler = "stan"
)

multi_nutpie <- fit_and_summarise(
  "multiphylo",
  data = authority$data,
  variables = list(
    political_authority = "ordered_logistic",
    religious_authority = "ordered_logistic"
  ),
  id = "language",
  tree = authority_multiphylo,
  prior = list(A_offdiag = "normal(0, 2)"),
  nuts_sampler = "nutpie",
  use_grad_based_mass_matrix=FALSE
)
```

## 5. Measurement error

```{r measurement-error}
set.seed(123)
n_me <- 20
tree_me <- rcoal(n_me)
d_me <- data.frame(
  id   = tree_me$tip.label,
  x    = rnorm(n_me),
  y    = rnorm(n_me),
  x_se = rexp(n_me, 5),
  y_se = rexp(n_me, 5)
)

me_stan <- fit_and_summarise(
  "measurement_error",
  data = d_me,
  variables = list(x = "normal", y = "normal"),
  id = "id",
  tree = tree_me,
  measurement_error = list(x = "x_se", y = "y_se"),
  nuts_sampler = "stan"
)

me_nutpie <- fit_and_summarise(
  "measurement_error",
  data = d_me,
  variables = list(x = "normal", y = "normal"),
  id = "id",
  tree = tree_me,
  measurement_error = list(x = "x_se", y = "y_se"),
  nuts_sampler = "nutpie",
  use_grad_based_mass_matrix=FALSE
)
```

## 6. Exact GP (spatial control)

```{r exact-gp}
gp_stan <- fit_and_summarise(
  "exact_gp",
  data = authority$data,
  variables = list(
    political_authority = "ordered_logistic",
    religious_authority = "ordered_logistic"
  ),
  id = "language",
  tree = authority$phylogeny,
  lon_lat = authority$coordinates,
  prior = list(A_offdiag = "normal(0, 2)"),
  nuts_sampler = "stan"
)

gp_nutpie <- fit_and_summarise(
  "exact_gp",
  data = authority$data,
  variables = list(
    political_authority = "ordered_logistic",
    religious_authority = "ordered_logistic"
  ),
  id = "language",
  tree = authority$phylogeny,
  lon_lat = authority$coordinates,
  prior = list(A_offdiag = "normal(0, 2)"),
  nuts_sampler = "nutpie",
  use_grad_based_mass_matrix=FALSE
)
```

## 7. HSGP (approximate GP, spatial control)

```{r hsgp}
hsgp_stan <- fit_and_summarise(
  "hsgp",
  data = authority$data,
  variables = list(
    political_authority = "ordered_logistic",
    religious_authority = "ordered_logistic"
  ),
  id = "language",
  tree = authority$phylogeny,
  lon_lat = authority$coordinates,
  dist_k = 5,
  prior = list(A_offdiag = "normal(0, 2)"),
  nuts_sampler = "stan"
)

hsgp_nutpie <- fit_and_summarise(
  "hsgp",
  data = authority$data,
  variables = list(
    political_authority = "ordered_logistic",
    religious_authority = "ordered_logistic"
  ),
  id = "language",
  tree = authority$phylogeny,
  lon_lat = authority$coordinates,
  dist_k = 5,
  prior = list(A_offdiag = "normal(0, 2)"),
  nuts_sampler = "nutpie",
  use_grad_based_mass_matrix=FALSE
)
```

## Combine and save results

```{r save-results}
all_ess <- bind_rows(
  auth_stan, auth_nutpie,
  prim_stan, prim_nutpie,
  rep_stan, rep_nutpie,
  multi_stan, multi_nutpie,
  me_stan, me_nutpie,
  gp_stan, gp_nutpie,
  hsgp_stan, hsgp_nutpie
)

saveRDS(all_ess, "nutpie_vs_stan_ess.rds")
write.csv(all_ess, "nutpie_vs_stan_ess.csv", row.names = FALSE)
```

## Results

### Wall-clock time

```{r plot-time, fig.width=8, fig.height=4}
time_df <- all_ess |>
  distinct(model, backend, elapsed)

ggplot(time_df, aes(x = model, y = elapsed, fill = backend)) +
  geom_col(position = "dodge") +
  labs(y = "Wall-clock time (seconds)", x = NULL, fill = "Backend") +
  scale_fill_manual(values = c(stan = "#4575b4", nutpie = "#d73027"))
```

### ESS distributions

```{r plot-ess-bulk, fig.width=9, fig.height=5}
ggplot(all_ess, aes(x = backend, y = ess_bulk, fill = backend)) +
  geom_violin(alpha = 0.6, draw_quantiles = c(0.25, 0.5, 0.75)) +
  facet_wrap(~model, scales = "free_y") +
  labs(y = "Bulk ESS", x = NULL) +
  scale_fill_manual(values = c(stan = "#4575b4", nutpie = "#d73027")) +
  theme(legend.position = "none")
```

```{r plot-ess-tail, fig.width=9, fig.height=5}
ggplot(all_ess, aes(x = backend, y = ess_tail, fill = backend)) +
  geom_violin(alpha = 0.6, draw_quantiles = c(0.25, 0.5, 0.75)) +
  facet_wrap(~model, scales = "free_y") +
  labs(y = "Tail ESS", x = NULL) +
  scale_fill_manual(values = c(stan = "#4575b4", nutpie = "#d73027")) +
  theme(legend.position = "none")
```

### ESS per second

```{r plot-ess-per-sec, fig.width=9, fig.height=5}
all_ess |>
  mutate(ess_bulk_per_sec = ess_bulk / elapsed) |>
  ggplot(aes(x = backend, y = ess_bulk_per_sec, fill = backend)) +
  geom_violin(alpha = 0.6, draw_quantiles = c(0.25, 0.5, 0.75)) +
  facet_wrap(~model, scales = "free_y") +
  labs(y = "Bulk ESS / second", x = NULL) +
  scale_fill_manual(values = c(stan = "#4575b4", nutpie = "#d73027")) +
  theme(legend.position = "none")
```

### Summary table

```{r summary-table}
all_ess |>
  group_by(model, backend) |>
  summarise(
    elapsed       = first(elapsed),
    median_ess    = median(ess_bulk, na.rm = TRUE),
    min_ess       = min(ess_bulk, na.rm = TRUE),
    median_ess_s  = median(ess_bulk / elapsed, na.rm = TRUE),
    max_rhat      = max(rhat, na.rm = TRUE),
    .groups = "drop"
  ) |>
  knitr::kable(digits = 1)
```
